Annotated example of a spatially varying effect with sdmTMB

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
minor edits to owls

Snow Owls are an arctic breeding species is an irruptive migrant meaning that the number of individuals choosing to migrate varies dramatically from year to year. 
The reasons for this variation are still not well understood, but seem to be at least partially related to very high abundances of food during the breeding season and therefore higher productivity [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

@@

```{r aic, echo=TRUE, cache=TRUE}
AIC(m0,m1,m2)
```

@@

Print the model to inspect the fit:
```{r check, include = FALSE, cache=FALSE, eval=TRUE}
m 
```

@@

```{r proj-p, echo=TRUE, eval=T}
p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

@@

```{r shapes, cache=FALSE}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=T, cache=FALSE}
b <- tidy(m)
p_mean <- p %>% group_by(CBCID) %>% summarise(
  X = mean(X), Y= mean(Y), 
  mean_est_count = exp(mean(est)), 
  nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's [Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

@@

```{r aic, echo=TRUE, cache=FALSE}
AIC(m0,m1,m2)
```

@@

Print the model to inspect the fit:
```{r check, include = FALSE, cache=FALSE, eval=TRUE}
m # loading this twice just prevents some excess messages from cluttering the output when printing below
```

@@

```{r proj-p}
p <- pred %>% mutate(X = X * 100000, Y = Y * 100000)
p_proj <- p %>% mutate(x = X, y = Y) %>% sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

@@

```{r shapes}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=FALSE}
b <- tidy(m)

p_mean <- p %>% group_by(CBCID) %>% summarise(
  X = mean(X), Y= mean(Y), 
  mean_est_count = exp(mean(est)), 
  nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=T, cache=T}
ggplot(data = p_proj) + geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = p_proj) + 
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=T}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```
